---
title: "Function Writing"
author: "Devin Pastoor"
output: pdf_document
---
## Writing Functions 

This section will be our bread and butter as functions provide the means for us to start to harness R's power to reduce duplication of code and increase our efficiency.

Functions in R are known as "first class objects" - that is, they can be treated like other R objects.

 They can be:
 
* created without a name
* assigned to variables
* stored in lists
* returned from functions
* passed as parameters to other functions

Essentially, you can do anything with a function that you can with a vector.

In R, a function is defined with the following syntax:

```
function(parameters) {body}
```
* `function` is a reserved word to initialize creation.
* **parameters** are sets of formal parameter names that will be defined in the function body.
* **Formal parameters** are parameters included in the function definition
* The **body** is simply the code that the function will execute

A function can be written in one line as shown above, however, to encapsulate multiple lines brackets `{}` must be used.

A multi-line function could look as such:

```
function(parameters) {
    some code
    some more code
    even more code
}
```

To create a function for future use we must assign it to an object (a variable). For example, we can a simple addition function to examine some features

```{r }
add_fun <- function(x, y) {
    x + y
}
```

This is a function declaration. We have created a function and given it a name. We can use it by calling it by name and passing some parameters that it requires.

```{r }
add_fun(1, 5)
```

There are numer of important behaviors going on 'behind-the-scenes' in even this simple function call.

### Default Behaviors

**Formal parameters** can be soley user defined, they can also have a default value/behavior.

Defaults can be assigned to a parameter with `=` 

Let's update our function to default to `y = 5`

```{r }
add_fun2 <- function(x, y = 5) x + y
```

When a default behavior defined, if no object or value is passed to that parameter, the default value is used.

```{r }
add_fun2(6)
```

But you can override the default behavior by simply passing in some value

```{r }
add_fun2(6, 3)
```

If no default is defined, the function will halt and give you an error requesting what to do with the missing parameter value

```{r, error = TRUE }
add_fun2(y = 3)
```

When you have multiple parameters - how does a function know which one to use for the various parameters?

### Parameter Matching

Like all things programming - R has specific rules for how it handles parameter matching for functions.

Here is a basic overview:

* parameters can be matched positionally or by name
* you can mix positional and named matching
    * when an parameter is matched by name it is "removed" from the parameter list - the remaining parameters are matched by order
* parameters can be partially matched

The overall order of operations for parameter matching:

1) Check exact match for named parameter
2) Check for partial match for named parameter
3) Check for positional match
4) Any remaining unmatched formal parameters are "taken up" by `...`

**Caveat(s)** 

* Any parameters *after* `...` are only matched exactly
* Tags partially matching multiple parameters will result in an error

### Passing on parameters

`...` parameter indicates that parameters may be passed on to other (internally called) functions

`...` can be used when extending another function where you don't want to copy all the parameters from the original function. 


```{r }
f <- function(x,...) {
    print(x)
    summary(...)
}

f("It worked! The summary is:", runif(1000, 0, 100), digits=2)
```

As you can see, all parameters after the first - which was given to `x` - where passed to `summary`.


### Return Values
In R after a function completes its code it will return a resulting value. 

```{r }
f <- function(x) x + 1
f(2)
```

`f(2)` returns the result of `x + 1`, which in this case is 3.

By default, **R returns the last evaluated expression**. You can also formally declare what you'd like R to return using `return()`

```{r }
f <- function(x) return(x + 1)
```

This can be helpful for legibility when dealing with more complex functions where multiple outcomes are possible. It can also help you "escape" a function early by returning a result as soon as one is relevant

```{r }
num_sign <- function(x) {
if (!is.numeric(x)) return("NaN")
if (x > 0) return("positive")
if (x <0) return("negative")
return("Don't know - is it zero?")
}

num_sign(1)
num_sign(-1)
num_sign("hello")
num_sign(0)

num_sign2 <- function(x) {
if (!is.numeric(x)) "NaN"
if (x > 0) "positive"
if (x <0) "negative"
"Don't know - is it zero?"
}

num_sign2(1)
num_sign2(-1)
num_sign2("hello")
num_sign2(0)
```

One other nomenclature change to note in the above example, is if an `if` statement
only contains a single statement on a single line, parentheses are not required. 

**R can only return a single result from a function**

To return multiple objects you can combine them into a list or other structure

```{r }
PK_info <- function(){
    id <- 1:10
    id
    doses <- c(1, 5, 10)
    doses
    time <- seq(0, 10, 1)
    time
}

PK_info2 <- function(){
    id <- 1:10
    doses <- c(1, 5, 10)
    time <- seq(0, 10, 1)
    list("id" = id, "doses" = doses, "time" = time)
}

PK_info() # only returns time
PK_info2() # returns everything as a list
```



## Types of Functions

3 specific types of functions that you may frequently run-into and/or utilize yourself are:

* Anonymous functions - functions that don't have a name

* Closures - functions written by other functions (will not discuss further)

* Lists of Functions - storing multiple functions in a list

### Anonymous Functions

In R, there is no special syntax for creating functions. Functions, like most things in R, are objects themselves. When you create a function, you are simply assigning a name to the object you are creating. By this behavior you can even create a function and assign it many names.

Sometimes, however, we don't want or need to spend the time assigning a name. You've most likely run across this when reading code that uses commands such as the `apply` family, `do.call`, or with `plyr`. 

Something along the lines of: 

`lapply(df, function(x) length(unique(x)))`

This lapply command could be rewritten with a named function

```
len_unique <- function(x) length(unique(x))
lapply(df, len_unique)
```

However, that is unnecessarily verbose for a one-time function, and can also introduce unnecessary clutter into your environment(s)

Just like other functions, anonymous function have **formals** (parameters), a **body** and are tied to a **parent environment**


### Lists of Functions
One way to store sets of functions is to put them in lists. Instead of a single function returning a list of results, you can actually store the functions themselves in a list for later re-use. 

```{r, eval = FALSE }
means <- list(
normal = function(x) mean(x),
geometric = function(x) ...,
harmonic = function(x) ...,
    )
```

To then call a function you can simply extract it from the list

`means$harmonic(data)` or `means[["geometric"]](data)`

While it might seem awkward there are some situations where lists of functions provide convenience. It also offers a degree of modularity that reduces dependencies.

#### Assignment

- Using a dataset of your choosing write 2-3 functions to create exploratory plots 
- Add those functions to an `exploratory_plots` list. 
- Using `lapply` quickly call all the functions on:
    - the whole dataset
    - a subset

