---
title: "Functional Programming"
author: "devin"
output: html_document
---

## Functional Programming

Functions provide a host of benefits to the user. The allow for efficient automation of repetitive tasks, bundling or common operations and a host of other possibilities.

One additional, and equally important, opporunity they offer is to **reduce errors**.

As touched upon in *Pragmatic Programming*, the **DRY** (DON'T REPEAT YOURSELF) is well suited to functions.

A Motivating Example

When dealing with a new dataset there are two frequent issues that arise with concentration data - BQL values, and how to handle them.

Given the hypothetical situation where you are given 3 similarly-structured datasets how you could handle replacing the phrase `LLOQ < 10` simply with NA?

The copy-paste process may be something along the lines of:

```
df1$conc[df1$CONC == 'LLOQ < 10'] <- NA
df2$conc[df2$CONC == 'LLOQ < 10'] <- NA
df3$conc[df3$CONC == 'LLOQ < 10'] <- NA
df4$conc[df4$CONC == 'LLOQ < 10'] <- NA
df5$conc[df4$CONC == 'LLOQ < 10'] <- NA
```

Quick - did anyone spot any issue with the code?

Let's write a function to help automate this, as well as reduce potential for mistakes such as the above


```{r eval = FALSE}
BQL_NA <- function(x) {
    x[x[["CONC"]] == 'LLOQ < 10'] <- NA
    x
}

BQL_NA(df1)
BQL_NA(df2)
BQL_NA(df3)
BQL_NA(df4)
BQL_MA(df5)
```

Closer, but again, we're repeating ourselves.

We touched on `lapply`, lets combine our dataframes into a list a 

**SIDE TRICK** `df[] <- lapply(df, our_fun)` - using `df[]` will give us back a dataframe instead of a list from `df <- lapply(df, our_fun)`

```{r, eval = FALSE }
df_list <- list(df1, df2, df3, df4, df5)
df_NOBQL <- lapply(df_list, BQL_NA)
```

### Assignment

- extend BQL_NA to allow us to pass different character strings for what how the LLOQ was defined
- BONUS: likewise, extend it to:   
    - handle any column 'conc' regardless of capitalization
    - handle any concentration column name (conc, concentration, DV, ...)

